Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-28 ms.tex
spelling

\citet{Ho:2009a} determines the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Much
evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaching the viscinity of the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities on galactic scales, which
temporarily increase the gas accretion rate towards the nucleus until
feedback becomes effective to shut off the flow and halt
SMBH growth.  Here we instead focus on time-independent models,
neglecting radiative cooling.  This approach allows us to
systematically explore the relevant parameter space and to derive
simple analytic expressions that prove useful in exploring whether
cooling instabilties manifest across the expected range of galaxy
properties, or whether other (non-AGN) forms of feedback can result in
steady accretion.  Even if cooling instabilties develop on galactic
scales over longer $\sim$ Gyr timescales, a state of
quasi-steady accretion may exist between these inflow events on
smaller radial scales comparable to the sphere of influence.  We are
thus motivated to explore whether steady flow is achieved within the
nucleus under any circumstances, or whether the gas content is always
building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

@@

\begin{table*}
\begin{threeparttable}
%\centering
\begin{minipage}{18cm}
\caption{Definitions of commonly used variables}
\begin{tabular}{ll}
\hline
{Variable} & {Definition} \\
\hline
$M_{\bullet}$ & Black hole mass \\
$\tilde{v}_{w}$ & Total heating parameter, including minimum heating
rate from stellar and black hole velocity dispersion \\
$v_{w}$ & Total heating parameter, excluding heating rate from velocity dispersion \\
$r_{s}$ & Stagnation radius (where gas radial velocity goes to zero) \\
$r_{\rm soi}$ & radius of sphere of influence (eq.~[\ref{eq:rsoi}]) \\
$r_{\rm b}$ & outer break radius of stellar density profile \\ 
$r_{\rm Ia}$ & radius interior to which Ia SN are infrequent compared to the dynamical/inflow timescale (eq.~[\ref{eq:rIa}]) \\ 
$\rho_{\star}(r)$ & 3D radial stellar density profile \\
$\rho(r)$ & gas density of CNM \\
$M_{\star}(r)$ & total enclosed stellar mass inside radius $r$ \\
$M_{\rm enc}(r)$ & total encolsed mass inside radius $r$ (SMBH + stars) \\
$q(r)$ & mass source term due to stellar winds (eq.~[\ref{eq:q}]) \\
$\eta$ & parameter setting normalization of mass input from stellar winds (eq.~[\ref{eq:q}]) \\
$\tau_{\star}$ & age of stellar population, in case of a single burst of star formation \\
$\Gamma$ & power-law slope of radial stellar density profile interior to the break radius \\
$\sigma(r)$ & stellar velocity dispersion \\
$\dot{M}$ & mass accretion rate \\
$\dot{M}_{\rm Ia}$ & maximum mass accretion rate as limited by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]) \\
$\dot{M}_{\rm C}$ & equilibrium accretion rate set by Compton heating (eq.~[\ref{eq:MdotC}]) \\
$\dot{M}_{\rm TI}$ & maximum accretion rate for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]) \\
$\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ & ratio of external heating (stellar winds, SN Ia, MSPs) to radiative cooling (eq.~[\ref{eq:cooling2}]) \\
$t_{\rm h}$ & Hubble timescale \\
$\nu$ & gas density power-law slope at the stagnation radius (eq.~[\ref{eq:densSlope}]) \\
\hline
\label{table:definitions}  
\end{tabular}
\begin{tablenotes}
\item{$^{(a)}FILL IN$}
\end{tablenotes}
\end{minipage}
\end{threeparttable}

@@

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\begin{equation}
\dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T),
\label{eq:qdotcond}
 \end{equation}
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

The gas density at the stagnation radius $\rho|_{r_{\rm s}}$ is more challenging to estimate accurately.  Using an alternative estimate of $\dot{M}$ as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}|_{r_{\rm s}} = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$,
\begin{align}
  &\dot{M}\sim\frac{(4 \pi/3) \rs^3 \rho|_{r_{\rm s}}}{\tff|_{r_{\rm s}}},
  \label{eq:mdot_gas}
\end{align}
 in conjuction with equations (\ref{eq:mdot_analytic}) and (\ref{eq:rs_simple}), one derives
\begin{align}
  \rho|_{r_{\rm s}}\approx
  \begin{cases}
    6 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core},\\
    1.0 \times 10^{-25}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}.
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

Our analytic derviations neglect the effects of heat conduction, an assumption we now check.  The ratio of the magnitude of the conductive heating rate (eq.~[\ref{eq:qdotcond}]) to the external heating rate at the stagnation radius is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}}
\tilde{v}_{w}^{2}}
 \left(\frac{1}{1+\frac{\kappa_0 T^{7/2}|_{r_{\rm s}}}{5 \rs
    \phi \rho c_s^3}}\right)
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  20 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 20 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  2 \phi & \text{saturated},
  \end{cases}
 \label{eq:conduction}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}),
(\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim
1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  We have approximated the
stellar density profile as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}|_{r_{\rm s}} &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    6.1 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    2.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
where the stagnation radius is assume to reside well inside the Nuker
break radius and for clarity we have not distinguished between cusp
and core galaxies.  Equation (\ref{eq:conduction}) shows that even in
the case of saturated conduction, the neglect of heat conduction
results in at most an order unity correction for plausible, causal values of
the saturation parameter $\phi < 0.1$, justifying its neglect in our
analysis.  Numerical experiments including conductivity confirm this
($\S\ref{sec:numerical}$).

@@

Fig.~\ref{fig:cooling} shows the ratio of wind heating to radiative
cooling, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$
(eq.~[\ref{eq:cooling2}] and surrounding discussion) as a function of
radius for our fiducial solutions shown in Figure \ref{fig:profiles}.
For high heating rates of $v_{w} = 600$ and $1200$ km s$^{-1}$ cooling
is unimportant across all radii, while for $v_{w} = 300$ km s$^{-1}$
we see that $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ and $\tcool/\tff$
can be less than unity, depending on the wind mass loss parameter
$\eta$.  At the stagnation radius, the ratio $\dot{q}_{\rm
  heat}/|\dot{q}_{\rm rad}|$ reaches a value that lies within a factor
of two of its minimum across the entire grid; thus, the value of
$(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ provides a
diagnostic of thermal instabilty for the entire solution.

@@

In Appendix \ref{app:windheat} we calculate the wind heating rate from
stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter,
$\eta$ (eq.~[\ref{eq:q}]), as a function of age, $\tau_{\star}$, of a
stellar population which is formed impulsively (Fig.~\ref{fig:vwImp}).  
At the earliest times ($\tau_{\star} \lesssim 10^{7}$ yr), the wind
heating rate exceeds 1000 km s$^{-1}$, while much later ($\tau_{\star}
\sim t_{\rm h}$) stellar wind heating dominated by main sequence winds
is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  We show
in $\S\ref{sec:combined}$ that for the case of quasi-continueous star
formation representative of the average star formation history of loss
mass galaxies, the stellar wind heating rate produced by stellar winds
can also be significant, $v_{\rm w}^{\star} \sim 1000$ km s$^{-1}$.  
Stellar winds thus contribute a potentially important source of both
energy and mass for the CNM.  An additional uncertainty in the stellar wind heating rate results from the uncertain thermalization efficiency of massive stellar winds.  We assume this efficiency to be unity, an optimistic assumption which is offset by our neglect of energy input from core collapse supernovae, which contribute an energy comparable to that from massive stellar winds.

@@

Based on the ATNF radio pulsar catalog (\citealt{Manchester+05}), we estimate the average spin-down luminosity of millisecond pulsars in the field to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$, resulting in $v_{w}^{\rm MSP} \lesssim 30$ km s$^{-1}$ for $\eta \gtrsim 0.02$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{w}^{\rm MSP}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the most optimistc assumptions $\epsilon_{\rm msp} = 1$ and $\eta = 0.02$.  On the other hand, it seems likely that the stellar binaries giving rise to MSPs could be disassociated by stellar interactions in the dense nuclear cluster, reducing the numbers of MSPs as compared to the field population estimate above.

@@

Using the results of \citet{Bromberg+11}, the time required for a jet of luminosity $L_{\rm
  j}$ and half opening angle $\theta_{\rm j} = 0.1$ (characteristic of
AGN jets) to propagate through a gaseous mass $M_{\rm g}$ of radius $r$ is estimated to be
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$, the ratio of the jet escape timescale to the dynamical timescale $t_{\rm dyn} \sim r/\sigma$ is given by
\begin{equation}
\frac{t_{\rm jet}}{t_{\rm dyn}} \sim 7\times 10^{-3}M_{\bullet,8}^{0.13}\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{-1/3},
\end{equation}
independent of $r$.  A jet with power sufficient to appreciably heat
the CNM on radial scales $\sim r_{\rm s}$ ($\epsilon_{j} \gtrsim
10^{-6}$) also necessarily has sufficient power to escape the nucler
region and propagate to much larger radii.  Slower outflows from the
accretion disk, instead of a collimated relativistic jet, may provide
a more promising source of feedback in these systems
(e.g.~\citealt{Li+13}).

@@

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the
accretion rate as a function of black hole mass for the average star
formation heating, calculated from equation (\ref{eq:eddr_analytic})
using our results for the total wind heating
(Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases
from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass
black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim
10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.
These fall below the maximum thermally stable accretion rate,
$\dot{M}_{\rm TI}$ (plotted as yellow lines), except at the highest
black hole masses $\sim 10^{9}M_{\odot}$.  For these highest mass
black holes, the maximum accretion rate as set by gas blow-out from SN
Ia (green lines; eq.~[\ref{eq:eddr_Ia}]) could instead cap the
accretion rate below the thermally-unstable value, provided that the
SN Ia heating efficiency is sufficiently high that $v_{\rm w}^{\rm Ia}
\gg \sqrt{2}\sigma \sim 400$ km s$^{-1}$ (see discussion just
preceeding eq.~[\ref{eq:eddr_Ia}]).  This is only possible if the Ia
heating efficiency is higher $\epsilon_{\rm Ia} > 0.1$ than normally
assumed (\citealt{Sharma+14}) and if the galaxy is a core.  For lower mass black holes, the
maximum accretion rate set by SN Ia is significantly higher and cannot
also prevent the development of thermally unstable flows,
i.e. $\dot{M}_{\rm Ia} \gg \dot{M}_{\rm TI}$.

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* calculated by \citet{Quataert:2004a}, as well as the range of $\dot{M}$ for the low-luminosity AGN NGC3115 from \citealt{ShcherbakovWong+:2014a} (who find a range of accretion rates depending on the assumed model for thermal conductivity, although they do not include saturation of the conductive flux).  The SgrA* rate exceeds our estimate of $\dot{M}$ due to the average star formation history for galaxies with the same mass black hole by almost two orders of magnitude.  This discrepency can be understood by the fact that star formation in the central parsec of the Galaxy is not in steady-state: feedback is instead dominated by the stellar winds from the ring of young massive stars of mass $\sim 10^{4}M_{\odot}$ and estimated age $\sim 10$ Myr (e.g., \citealt{Schodel+07}).  Such a star formation history is better described as a `burst' of stars, for which the value of $\eta \sim 100$ at $\tau_{\star} = 10^{7}$ yr (Fig.~\ref{fig:vwImp}) is a factor of $\sim 100$ times higher than the value of $\eta \sim 1$ predicted in the average SFH case (Fig.~\ref{fig:vwSources}, bottom panel).

>>>>

\citet{Ho:2009a} determines the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Much
evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaching the vicinity of the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities on galactic scales, which
temporarily increase the gas accretion rate towards the nucleus until
feedback becomes effective to shut off the flow and halt
SMBH growth.  Here we instead focus on time-independent models,
neglecting radiative cooling.  This approach allows us to
systematically explore the relevant parameter space and to derive
simple analytic expressions that prove useful in exploring whether
cooling instabilities manifest across the expected range of galaxy
properties, or whether other (non-AGN) forms of feedback can result in
steady accretion.  Even if cooling instabilities develop on galactic
scales over longer $\sim$ Gyr timescales, a state of
quasi-steady accretion may exist between these inflow events on
smaller radial scales comparable to the sphere of influence.  We are
thus motivated to explore whether steady flow is achieved within the
nucleus under any circumstances, or whether the gas content is always
building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

@@

\begin{table*}
\begin{threeparttable}
%\centering
\begin{minipage}{18cm}
\caption{Definitions of commonly used variables}
\begin{tabular}{ll}
\hline
{Variable} & {Definition} \\
\hline
$M_{\bullet}$ & Black hole mass \\
$\tilde{v}_{w}$ & Total heating parameter, including minimum heating
rate from stellar and black hole velocity dispersion \\
$v_{w}$ & Total heating parameter, excluding heating rate from velocity dispersion \\
$r_{s}$ & Stagnation radius (where gas radial velocity goes to zero) \\
$r_{\rm soi}$ & radius of sphere of influence (eq.~[\ref{eq:rsoi}]) \\
$r_{\rm b}$ & outer break radius of stellar density profile \\ 
$r_{\rm Ia}$ & radius interior to which Ia SN are infrequent compared to the dynamical/inflow timescale (eq.~[\ref{eq:rIa}]) \\ 
$\rho_{\star}(r)$ & 3D radial stellar density profile \\
$\rho(r)$ & gas density of CNM \\
$M_{\star}(r)$ & total enclosed stellar mass inside radius $r$ \\
$M_{\rm enc}(r)$ & total enclosed mass inside radius $r$ (SMBH + stars) \\
$q(r)$ & mass source term due to stellar winds (eq.~[\ref{eq:q}]) \\
$\eta$ & parameter setting normalization of mass input from stellar winds (eq.~[\ref{eq:q}]) \\
$\tau_{\star}$ & age of stellar population, in case of a single burst of star formation \\
$\Gamma$ & power-law slope of radial stellar density profile interior to the break radius \\
$\sigma(r)$ & stellar velocity dispersion \\
$\dot{M}$ & mass accretion rate \\
$\dot{M}_{\rm Ia}$ & maximum mass accretion rate as limited by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]) \\
$\dot{M}_{\rm C}$ & equilibrium accretion rate set by Compton heating (eq.~[\ref{eq:MdotC}]) \\
$\dot{M}_{\rm TI}$ & maximum accretion rate for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]) \\
$\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ & ratio of external heating (stellar winds, SN Ia, MSPs) to radiative cooling (eq.~[\ref{eq:cooling2}]) \\
$t_{\rm h}$ & Hubble timescale \\
$\nu$ & gas density power-law slope at the stagnation radius (eq.~[\ref{eq:densSlope}]) \\
\hline
\label{table:definitions}  
\end{tabular}
\begin{tablenotes}
\item{$^{(a)}FILL IN$}
\end{tablenotes}
\end{minipage}
\end{threeparttable}

@@

To isolate the physics of interest, our baseline calculations neglects three potentially important effects: heat conduction, radiative cooling, and rotation.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\begin{equation}
\dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T),
\label{eq:qdotcond}
 \end{equation}
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conductivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity errors in the key properties of the solutions.

@@

The gas density at the stagnation radius $\rho|_{r_{\rm s}}$ is more challenging to estimate accurately.  Using an alternative estimate of $\dot{M}$ as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}|_{r_{\rm s}} = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$,
\begin{align}
  &\dot{M}\sim\frac{(4 \pi/3) \rs^3 \rho|_{r_{\rm s}}}{\tff|_{r_{\rm s}}},
  \label{eq:mdot_gas}
\end{align}
 in conjunction with equations (\ref{eq:mdot_analytic}) and (\ref{eq:rs_simple}), one derives
\begin{align}
  \rho|_{r_{\rm s}}\approx
  \begin{cases}
    6 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core},\\
    1.0 \times 10^{-25}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}.
  \end{cases}
  \label{eq:rhors}
\end{align}

@@

Our analytic derivations neglect the effects of heat conduction, an assumption we now check.  The ratio of the magnitude of the conductive heating rate (eq.~[\ref{eq:qdotcond}]) to the external heating rate at the stagnation radius is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}}
\tilde{v}_{w}^{2}}
 \left(\frac{1}{1+\frac{\kappa_0 T^{7/2}|_{r_{\rm s}}}{5 \rs
    \phi \rho c_s^3}}\right)
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  20 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 20 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  2 \phi & \text{saturated},
  \end{cases}
 \label{eq:conduction}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}),
(\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim
1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  We have approximated the
stellar density profile as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}|_{r_{\rm s}} &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    6.1 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    2.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
where the stagnation radius is assume to reside well inside the Nuker
break radius and for clarity we have not distinguished between cusp
and core galaxies.  Equation (\ref{eq:conduction}) shows that even in
the case of saturated conduction, the neglect of heat conduction
results in at most an order unity correction for plausible, causal values of
the saturation parameter $\phi < 0.1$, justifying its neglect in our
analysis.  Numerical experiments including conductivity confirm this
($\S\ref{sec:numerical}$).

@@

Fig.~\ref{fig:cooling} shows the ratio of wind heating to radiative
cooling, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$
(eq.~[\ref{eq:cooling2}] and surrounding discussion) as a function of
radius for our fiducial solutions shown in Figure \ref{fig:profiles}.
For high heating rates of $v_{w} = 600$ and $1200$ km s$^{-1}$ cooling
is unimportant across all radii, while for $v_{w} = 300$ km s$^{-1}$
we see that $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ and $\tcool/\tff$
can be less than unity, depending on the wind mass loss parameter
$\eta$.  At the stagnation radius, the ratio $\dot{q}_{\rm
  heat}/|\dot{q}_{\rm rad}|$ reaches a value that lies within a factor
of two of its minimum across the entire grid; thus, the value of
$(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}}$ provides a
diagnostic of thermal instability for the entire solution.

@@

In Appendix \ref{app:windheat} we calculate the wind heating rate from
stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter,
$\eta$ (eq.~[\ref{eq:q}]), as a function of age, $\tau_{\star}$, of a
stellar population which is formed impulsively (Fig.~\ref{fig:vwImp}).  
At the earliest times ($\tau_{\star} \lesssim 10^{7}$ yr), the wind
heating rate exceeds 1000 km s$^{-1}$, while much later ($\tau_{\star}
\sim t_{\rm h}$) stellar wind heating dominated by main sequence winds
is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  We show
in $\S\ref{sec:combined}$ that for the case of quasi-continuous star
formation representative of the average star formation history of loss
mass galaxies, the stellar wind heating rate produced by stellar winds
can also be significant, $v_{\rm w}^{\star} \sim 1000$ km s$^{-1}$.  
Stellar winds thus contribute a potentially important source of both
energy and mass for the CNM.  An additional uncertainty in the stellar wind heating rate results from the uncertain thermalization efficiency of massive stellar winds.  We assume this efficiency to be unity, an optimistic assumption which is offset by our neglect of energy input from core collapse supernovae, which contribute an energy comparable to that from massive stellar winds.

@@

Based on the ATNF radio pulsar catalog (\citealt{Manchester+05}), we estimate the average spin-down luminosity of millisecond pulsars in the field to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$, resulting in $v_{w}^{\rm MSP} \lesssim 30$ km s$^{-1}$ for $\eta \gtrsim 0.02$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{w}^{\rm MSP}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the most optimistic assumptions $\epsilon_{\rm msp} = 1$ and $\eta = 0.02$.  On the other hand, it seems likely that the stellar binaries giving rise to MSPs could be disassociated by stellar interactions in the dense nuclear cluster, reducing the numbers of MSPs as compared to the field population estimate above.

@@

Using the results of \citet{Bromberg+11}, the time required for a jet of luminosity $L_{\rm
  j}$ and half opening angle $\theta_{\rm j} = 0.1$ (characteristic of
AGN jets) to propagate through a gaseous mass $M_{\rm g}$ of radius $r$ is estimated to be
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$, the ratio of the jet escape timescale to the dynamical timescale $t_{\rm dyn} \sim r/\sigma$ is given by
\begin{equation}
\frac{t_{\rm jet}}{t_{\rm dyn}} \sim 7\times 10^{-3}M_{\bullet,8}^{0.13}\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{-1/3},
\end{equation}
independent of $r$.  A jet with power sufficient to appreciably heat
the CNM on radial scales $\sim r_{\rm s}$ ($\epsilon_{j} \gtrsim
10^{-6}$) also necessarily has sufficient power to escape the nuclear
region and propagate to much larger radii.  Slower outflows from the
accretion disk, instead of a collimated relativistic jet, may provide
a more promising source of feedback in these systems
(e.g.~\citealt{Li+13}).

@@

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the
accretion rate as a function of black hole mass for the average star
formation heating, calculated from equation (\ref{eq:eddr_analytic})
using our results for the total wind heating
(Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases
from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass
black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim
10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.
These fall below the maximum thermally stable accretion rate,
$\dot{M}_{\rm TI}$ (plotted as yellow lines), except at the highest
black hole masses $\sim 10^{9}M_{\odot}$.  For these highest mass
black holes, the maximum accretion rate as set by gas blow-out from SN
Ia (green lines; eq.~[\ref{eq:eddr_Ia}]) could instead cap the
accretion rate below the thermally-unstable value, provided that the
SN Ia heating efficiency is sufficiently high that $v_{\rm w}^{\rm Ia}
\gg \sqrt{2}\sigma \sim 400$ km s$^{-1}$ (see discussion just
preceding eq.~[\ref{eq:eddr_Ia}]).  This is only possible if the Ia
heating efficiency is higher $\epsilon_{\rm Ia} > 0.1$ than normally
assumed (\citealt{Sharma+14}) and if the galaxy is a core.  For lower mass black holes, the
maximum accretion rate set by SN Ia is significantly higher and cannot
also prevent the development of thermally unstable flows,
i.e. $\dot{M}_{\rm Ia} \gg \dot{M}_{\rm TI}$.

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* calculated by \citet{Quataert:2004a}, as well as the range of $\dot{M}$ for the low-luminosity AGN NGC3115 from \citealt{ShcherbakovWong+:2014a} (who find a range of accretion rates depending on the assumed model for thermal conductivity, although they do not include saturation of the conductive flux).  The SgrA* rate exceeds our estimate of $\dot{M}$ due to the average star formation history for galaxies with the same mass black hole by almost two orders of magnitude.  This discrepancy can be understood by the fact that star formation in the central parsec of the Galaxy is not in steady-state: feedback is instead dominated by the stellar winds from the ring of young massive stars of mass $\sim 10^{4}M_{\odot}$ and estimated age $\sim 10$ Myr (e.g., \citealt{Schodel+07}).  Such a star formation history is better described as a `burst' of stars, for which the value of $\eta \sim 100$ at $\tau_{\star} = 10^{7}$ yr (Fig.~\ref{fig:vwImp}) is a factor of $\sim 100$ times higher than the value of $\eta \sim 1$ predicted in the average SFH case (Fig.~\ref{fig:vwSources}, bottom panel).
